---
title: "Kick the Bar Chart Habit: Boxplots"
teaching: 60
exercises: 30
questions:
- "What are the limitations of bar charts?"
- "What information do boxplots display?"
- "How can I subset data ?"
objectives:
- "Load data from a URL."
- "Describe the anatomy of a box plot."
- "Use box plots to compare mean and median values between groups."
- "Order groups by mean value."
- "Subset data."
- "Describe some pitfalls of using bar charts to display or compare means."
keypoints:
- ""
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("02-")
```

```{r, include=FALSE}
library(raster)
```

## Preliminaries
Bar charts are useful for displaying count data but are often used to portray statistical information that they don't represent well. In this lesson we'll learn to ['Kick the bar chart habit'](http://www.nature.com/nmeth/journal/v11/n2/full/nmeth.2837.html) by creating box plots as an alternative to bar charts. This lesson uses data from a multi-system survey of mouse physiology in 8 inbred founder strains and 54 F1 hybrids of the Collaborative Cross. The study is described in [Lenarcic et al, 2012](http://www.genetics.org/content/190/2/413.full). For more information about this data set, see the [CGDpheno3 data](http://phenome.jax.org/projects/CGDpheno3) at Mouse Phenome Database. 

#### Load package and library
Load the ggplot library. You'll need to install the packages
first if you haven't done so already. Install them from the `Packages` tab in RStudio, or use the `install.packages()` command in the Console. Use double quotes around the package name.

```{r eval = FALSE}
install.packages("ggplot2")
```

You only need to install a package once to download it into your machine's library. Once you have installed the package on your machine, you need to load the library into R in order to use the functions contained in the package.  

```{r}
library(ggplot2)
```

#### Load and explore data
Load the data from this shortened URL. Mind the double quotes.  

```{r}
cc_data <- read.csv(file = "http://bit.ly/CGDPheno3")
```

Explore the data variables. The first 4 columns contain strain, sex, and ID numbers. The remaining contain phenotype measurements with abbreviated names.

```{r}
names(cc_data)
```

How many mice? 
```{r}
dim(cc_data)
```

There are `r nrow(cc_data)` mice in rows, and `r ncol(cc_data)` columns containing strain, sex, ID numbers, and phenotypes.

How many mice of each sex? 

```{r}
table(cc_data$sex)
```

How many mice of each strain? 

```{r}
table(cc_data$strain)
```

In most cases there are `r modal(table(cc_data$strain))` mice of each strain, with some exceptions. Some strains only have only `r min(table(cc_data$strain))` mice, while others have as many as `r max(table(cc_data$strain))`.  
How many mice of each strain by sex?

```{r}
table(cc_data$sex, cc_data$strain)
```

Most strains have `r modal(table(cc_data$sex, cc_data$strain))` mice of each sex, while some may have `r min(table(cc_data$sex, cc_data$strain))` or as many as `r max(table(cc_data$sex, cc_data$strain))` of one sex. How do the first few rows of data look? Note the NAs in the data. These are missing values and can complicate analyses unless specifically addressed. 

```{r}
head(cc_data)
```

## Plotting with [ggplot](http://ggplot2.org/)
Use the `ggplot()` function, which is found in the `ggplot2` library. The basic `ggplot()` syntax is: `ggplot(data, mapping) + layer()`. We will build a plot of red blood cells by strain in several steps, addressing many of the ways that you can fine-tune your plot to display exactly the information that you want in the way that you want to do so. 

```{r fig.height=10, fig.width=10, warning=FALSE}
ggplot(data = cc_data, mapping = aes(x = strain, y = RBC)) + 
  geom_boxplot()
```

In a boxplot, the upper whisker extends to the highest value within 1.5 x inter-quartile range (IQR, or distance between first and third quartiles) and the lower whisker extends to the lowest value within 1.5 x IQR of the hinge. Data beyond the end of the whiskers (outliers) are plotted as points.

You can view summary statistical information about a phenotype, such as the first quartile, median, or third quartile values by using `summary()`.

```{r}
summary(object = cc_data$RBC)
```

The median red blood cell count across all strains is `r median(cc_data$RBC, na.rm=TRUE)` and the mean value is `r mean(cc_data$RBC, na.rm=TRUE)`. If you wanted summary statistics for a specific strain, you could subset the data by specifying the strain name.

```{r}
summary(cc_data$RBC[cc_data$strain=="NOD/ShiLtJ"])
```

We'll look at subsetting data more in a bit. Back to plotting.

It's difficult to distinguish the strain names on the x-axis, so flip the coordinates to place strain on the y-axis and red blood cells on the x-axis. 

```{r fig.height=10, fig.width=10, warning=FALSE}
ggplot(data = cc_data, mapping = aes(x = strain, y = RBC)) + 
  geom_boxplot() + 
  coord_flip()
```

Sort the strains by mean red blood cells. Do this by re-ordering strains within the mapping function `aes()`. Save the plot as a variable.

```{r fig.height=10, fig.width=10, warning=FALSE}
ggplot(data = cc_data, mapping = aes(x = reorder(strain, RBC, FUN = "mean", na.rm = TRUE), y = RBC)) + 
  geom_boxplot() + 
  coord_flip()
```

Add a point indicating the mean RBC value for each strain. Add a statistical summary layer to do this.

```{r fig.height=10, fig.width=10, warning=FALSE}
ggplot(data = cc_data, mapping = aes(x = reorder(strain, RBC, FUN = "mean", na.rm = TRUE), y = RBC)) + 
  geom_boxplot() + 
  coord_flip() + 
  stat_summary(fun.y = "mean", geom = "point")
```

You should see a purple square indicating the mean red blood cell value for each strain. Is it the same as the median value for each strain?
Notice that the mean value is sensitive to outliers, while the median value is not sensitive to outliers. 
Find the boxplot for WSBCASTF1. Notice that a single data point with a value greater than 11.5 pulls the mean value for this strain far over to the right.

Plot the data points over each boxplot. Since ggplot builds a plot layer by layer, the boxplot layer should come before the data points so as not to obscure them.

```{r fig.height=10, fig.width=10, warning=FALSE}
ggplot(data = cc_data, mapping = aes(x = reorder(strain, RBC, FUN = "mean", na.rm = TRUE), y = RBC)) + 
  geom_boxplot() + 
  geom_point() +
  coord_flip() + 
  stat_summary(fun.y = "mean", geom = "point")
```

  Color the data points by sex. Save the plot as a variable. To view the plot, type the name of the variable.  

```{r fig.height=10, fig.width=10, warning=FALSE}
rbc_boxplot <- ggplot(data = cc_data, mapping = aes(x = reorder(strain, RBC, FUN = "mean", na.rm = TRUE),
                                                    y = RBC)) + 
  geom_boxplot() + 
  geom_point(aes(colour = sex)) +
  coord_flip() + 
  stat_summary(fun.y = "mean", geom = "point")
rbc_boxplot
```

  Add axis labels. Redefine the plot variable.  

```{r fig.height=10, fig.width=10, warning=FALSE}
rbc_boxplot <- rbc_boxplot +
  xlab("strain") + 
  ylab("red blood cell count (n/uL)")
rbc_boxplot
```

  Add a title. Redefine the plot variable.  

```{r fig.height=10, fig.width=10, warning=FALSE}
rbc_boxplot <- rbc_boxplot + 
  ggtitle("Red Blood Cell Distribution by Strain")
rbc_boxplot
```

> ## Challenge 1
> 1. Choose another phenotype to plot as boxplots by strain.
> 2. Flip the coordinates if necessary to make strain names legible.
> 3. Order boxplots by mean phenotype value.
> 4. Add a point indicating the mean strain value. 
> 5. Add data points over the boxplots (optional). 
> 6. Add axis labels and a plot title.
> > ## Solution to Challenge 1
> > 1. For percent neutrophils: `ggplot(data = cc_data, mapping = aes(x = strain, y = pctNEUT)) + geom_boxplot()`
> > 2. For percent neutrophils: `ggplot(data = cc_data, mapping = aes(x = strain, y = pctNEUT)) + geom_boxplot() + coord_flip()`
> > 3. `ggplot(data = cc_data, mapping = aes(x = reorder(strain, pctNEUT, FUN = "mean", na.rm = TRUE), y = pctNEUT)) + geom_boxplot() + coord_flip()`
> > 4. `ggplot(data = cc_data, mapping = aes(x = reorder(strain, pctNEUT, FUN = "mean", na.rm = TRUE), y = pctNEUT)) + geom_boxplot() + coord_flip() + stat_summary(fun.y = "mean", geom = "point")`
> > 5. `ggplot(data = cc_data, mapping = aes(x = reorder(strain, pctNEUT, FUN = "mean", na.rm = TRUE), y = pctNEUT)) + geom_boxplot() + geom_point() + coord_flip() + stat_summary(fun.y = "mean", geom = "point")`
> > 6. `ggplot(data = cc_data, mapping = aes(x = reorder(strain, pctNEUT, FUN = "mean", na.rm = TRUE), y = pctNEUT)) + geom_boxplot() + geom_point() + coord_flip() + stat_summary(fun.y = "mean", geom = "point")  + xlab("strain") + ylab("percent neutrophils") + ggtitle("Percent Neutrophils by Strain")`

#### Subsetting data
Select a subset of the strains. Choose strains with the highest and lowest mean and median red blood cell counts.
Include the parent strains of the F1s.  

```{r fig.height=10, fig.width=10, warning=FALSE}
subset.cc_data <- subset(cc_data, strain %in% c("ACASTF1", "APWKF1", "CAST/EiJ", "PWK/PhJ",
                                      "A/J", "NODAF1", "NOD/ShiLtJ") == TRUE)
```

  Create boxplots from the subset.  

```{r fig.height=10, fig.width=10, warning=FALSE}
ggplot(data = subset.cc_data, mapping = aes(x = strain, y = RBC)) + 
  geom_boxplot()
```

Order by mean RBC value as before. Save the plot as a variable.

```{r fig.height=10, fig.width=10, warning=FALSE}
subset_boxplot <- ggplot(data = subset.cc_data,
                         mapping = aes(x = reorder(strain, RBC, FUN = "mean", na.rm = TRUE),
                                       y = RBC)) + 
  geom_boxplot()
subset_boxplot
```

This time there's no need to flip the axes since the strain names are legible on the x-axis. 
Plot the data points by sex.The boxplots have already been drawn and saved in the variable `subset_boxplot`. Layer the data points on top of the boxplots.

```{r fig.height=10, fig.width=10, warning=FALSE}
subset_boxplot <- subset_boxplot + 
  geom_point(aes(colour = sex))
subset_boxplot
```

Add a purple square indicating the mean RBC value for each strain.

```{r fig.height=10, fig.width=10, warning=FALSE}
subset_boxplot <- subset_boxplot + 
  stat_summary(fun.y = "mean", geom = "point", colour = "mediumpurple4", shape = 15, size = 2)
subset_boxplot
```

Add x and y axis labels.

```{r fig.height=10, fig.width=10, warning=FALSE}
subset_boxplot <- subset_boxplot + 
  xlab("strain") + 
  ylab("red blood cell count (n/uL)")
subset_boxplot
```

Add a title.

```{r fig.height=10, fig.width=10, warning=FALSE}
subset_boxplot <- subset_boxplot + 
  ggtitle("Red Blood Cell Distribution by Strain")
subset_boxplot
```

Output the plot to a PDF file. Set width and height. Turn off the output to pdf with the `dev.off()` command.

```{r warning=FALSE}
pdf("subset-boxplot.pdf", width= 8, height = 9)
print(subset_boxplot)
dev.off()
```

Picture a set of bar charts indicating the mean and s.e.m. for each strain. Which plot communicates more information, a bar chart or a box plot?




## References
1. Kick the bar chart habit. Nat Meth. 2014;11(2):113. doi: 10.1038/nmeth.2837. 
2. Lenarcic AB, Svenson KL, Churchill GA, Valdar W. A general Bayesian approach to analyzing diallel crosses of inbred strains. Genetics. 2012 Feb 1;190(2):413-35.
3. Spitzer M, Wildenhain J, Rappsilber J, Tyers M. BoxPlotR: a web tool for generation of box plots. Nat Meth. 2014;11(2):121-2. doi: 10.1038/nmeth.2811.
4. Krzywinski M, Altman N. Points of Significance: Visualizing samples with box plots. Nat Meth. 2014;11(2):119-20. doi: 10.1038/nmeth.2813.
